---
title: "Who's the Victim in a US-China Trade War?"
author: "Authored by Stephen Stapleton"
date: "February 9, 2019"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(plyr)
library(dplyr)
library(ggplot2)
library(zoo)
library(here)
library(readxl)
library(ggrepel)
library(scales)
library(extrafont)
library(sf)

# font_import()
# loadfonts(device = 'win')

# Set my custom theme for project
theme_staples <- theme_bw(base_size = 14) +
  theme(
    text=element_text(family = "Saira Condensed"),
    plot.title =element_text(family = "Archivo Narrow"),
    axis.line = element_line(color = "grey30"),
    axis.title = element_text(color = "grey30"),
    legend.background = element_rect(fill=NA, color = NA),
    legend.key = element_rect(fill=NA, color=NA),
    legend.title = element_text(color = "grey30"),
    legend.text = element_text(color = "grey30"),
    panel.background  = element_blank(),
    panel.border = element_blank(),
    plot.background = element_rect(fill=NA, colour=NA),
    panel.grid = element_line(color = "grey95")
  )

theme_set(theme_staples)

# Set my main colors
staples_colors <- c(
  `red1` = "#CC0032",
  `blue1` = "#149DCC",
  `teal1` = "#00FFB5",
  `red2` = "#FF5A4C",
  `blue2` = "#317E99"
  )

# Function to extract cols as hex codes
# Adapted from https://drsimonj.svbtle.com/creating-corporate-colour-palettes-for-ggplot2
# Call cols e.g. staples_cols("red")
staples_cols <- function(...) {
  cols <- c(...)
  if (is.null(cols))
    return (staples_colors)
  staples_colors[cols]
}

# Define my palettes
staples_palettes <- list(
  `main` = staples_cols("red1", "red2", "teal1", "blue1", "blue2"),
  `cool` = staples_cols("blue2", "teal1"),
  `hot` = staples_cols("teal1", "red1"),
  `light` = staples_cols('red2', 'blue1')
)

# Return function interpolating color palettes
staples_pal <- function(palette = "main", reverse = FALSE, ...) {
  pal <- staples_palettes[[palette]]
  if (reverse) pal <- rev(pal)
  colorRampPalette(pal, ...)
}

# Color scale constructor for cols
scale_color_staples <- function(palette = "main", discrete = TRUE, reverse = FALSE, ...) {
  pal <- staples_pal(palette = palette, reverse = reverse)
  if (discrete) {
    discrete_scale("colour", paste0("staples_", palette), palette = pal, ...)
  } else {
    scale_color_gradientn(colours = pal(256), ...)
  }
}

# Fill scale constructor for cols
scale_fill_staples <- function(palette = "main", discrete = TRUE, reverse = FALSE, ...) {
  pal <- staples_pal(palette = palette, reverse = reverse)
  if (discrete) {
    discrete_scale("fill", paste0("staples_", palette), palette = pal, ...)
  } else {
    scale_fill_gradientn(colours = pal(256), ...)
  }
}

```

## Introduction

Trade wars often captivate the media and broader public discourse, with flights of coverage going through the chaotic highs and lows of retaliatory and conciliatory action. In such moments, it is often difficult to decouple the national drama from the facts of what a trade war truly is, and, perhaps more importantly, who is left paying the tab when a new trade equilibrium is reached. This project is intended to provide a short respite from the back and forth, and to go into the specifics of what a trade war means for the American people. My mission is to present the hard numbers, and to give some of the best estimates as to what changes in the global market we may expect, and what regions are likely to be hit hardest with job layoffs and rising costs.

## First, why China?

The question is a legitimate one, and is the best place to start when grasping the scope of effect from a trade war. When the Chinese market first opened up in earnest to the world, it provided an opportunity for many companies looking for cheap labor to make a substantial profit by producing cheap and selling to developed nation consumers. Since then, trade between China and the World has exploded, lifting billions out of poverty in China and enabling substantial drops in the price of most consumer goods in the United States. As such, demand in the United States for cheaper Chinese goods has led to greater and greater import levels.  The United States now runs the largest trade deficit in the World with China, since as a nation we buy substantially more than we sell. Part of this growing trend was China's entry into the World Trade Organization in 2001, which eliminated many of the trade barriers that restricted free trade. While weighting our trade balances by gross domestic product (GDP) indicates this imbalance is higher per capita in many other nations, policymakers nevertheless have expressed concern that without intervention, this imbalance will continue to widen.

```{r fig.width = 9, warning = F, message = F, dpi=300}
# Source: https://www.census.gov/foreign-trade/balance/index.html
USall <- read_csv('country.csv')

# Parse dates for use in ggplot
USall$new_date <- as.Date(as.yearmon(USall$date))

# Redefine variable deficit to trade balance
USall$trd_bal <- -USall$deficit

# Restrict to years 1985-2017 for yearly total
USall <- USall[USall$new_date < '2018-09-01',]

# Exclude regional trade comparisons
USuni <- USall[USall$cty_code > 1000,]

# Summarize over year to adjust for seasonal spikes
USyr <- ddply(USuni, .(cty_code, year), summarize, trd_bal = sum(trd_bal))

# Prepare for ggplot
USyr$is_china <- ifelse(USyr$cty_code == 5700, 1, 0)
USyr$trd_bal <- USyr$trd_bal/1000
USyr <- USyr[USyr$year < 2018,]

# Plot trade balance per country with coloring and alpha aesthetics
# Create non-aesthetic xintercept for WTO adoption by China
# Employ geom_line(), geom_vline()
# Adjust scaling for both x and y axis
ggplot(USyr, aes(x = year, y = trd_bal,
  color = as.factor(cty_code), alpha = as.factor(is_china))) +
  geom_line() +
  geom_vline(xintercept = 2001, linetype = 'dashed', color = staples_cols("teal1")) +
  scale_alpha_manual(values = c(.4, 1)) +
  scale_color_manual(values = c(rep('grey', 150), '#CC2414', rep('grey', 93))) +
  scale_x_continuous(breaks = seq(1985, 2017, 5), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(-400, 100, 100), expand = c(0,0)) +
  annotate("text", x = 2015, y = -310, label = 'China', color = staples_cols("red1"), family = "Saira Condensed", size = 5) +
  annotate("text", x = 2000.5, y = -200, label = 'China Enters WTO', angle = 90, color = "grey30", family = "Saira Condensed", size = 5) +
  annotate("rect", xmin = 2008, xmax = 2009, ymin = -400, ymax = 100, alpha = .1, fill = "grey70") +
  theme(legend.position = 'none',
        panel.grid = element_blank()) +
  labs(x = 'YEAR', y = 'TRADE BALANCE (BILLIONS $USD)',
       title = 'The US-China trade gap stands out among US trade relationships',
       subtitle = 'In nominal terms, the deficit trends downward after China entered the World Trade Organization in December 2001, \nquadrupling since 2000. While population weighting reduces the deficit somewhat, it does not change this trend.', 
       caption = 'Data Source: United States Census Bureau')
```

Breaking the trade relationship down to the state level, China runs a surplus with many of its primary state trading partners. A balance-trade line shows most states having China as their primary foreign trade partner run a deficit, with Washington State a notable outlier. Many states rely substantially on trade with China as a part of their local economy. California, Illinois, Washington, Georgia, and Tennessee each attribute roughly five to eight percent of their state GDP to trade with China (imports and exports), whereas Louisiana alone runs a substantive surplus. These states are then ones with particularly more to lose from a US-China trade war, and will be explored in greater detail as this project continues.

``` {r fig.width = 9, warning = F, message = F, dpi=300}
# source: https://www.census.gov/foreign-trade/statistics/state/data/index.html
EXall <- read_excel('exctyall.xls',
            range = cell_rows(5:1436),
            col_types = c(rep('text', 3), rep('numeric', 9))
            )
IMall <- read_excel('imctyall.xls',
            range = cell_rows(5:1436),
            col_types = c(rep('text', 3), rep('numeric', 9))
            )

# extract only 2017 imports/exports
IMall$IMval2017 <- IMall$val2017
EXall$EXval2017 <- EXall$val2017

IMall$IMshr2017 <- IMall$share17
EXall$EXshr2017 <- EXall$share17

IMall$IMrank <- IMall$rank
EXall$EXrank <- EXall$rank

# combine by state and trading country
ImEx <- left_join(IMall, EXall, by = c('statename', 'countryd'))

# subset only chinese imports and exports
ImExCN <- subset(ImEx, (ImEx$countryd %in% 'China'))

# specify if china is top trading partner
ImExCN$trd_main <- ifelse(ImExCN$IMrank %in% c(1:3) & ImExCN$EXrank %in% c(1:3), 'Yes', 'No')

# convert millions to billions for each of graph reading
ImExCN$IMval2017 <- ImExCN$IMval2017/1000
ImExCN$EXval2017 <- ImExCN$EXval2017/1000

ImExCN$IMshr2017 <- ImExCN$IMshr2017/100
ImExCN$EXshr2017 <- ImExCN$EXshr2017/100

# Create relative measure of (imports+exports)/gdp per state
# Source: https://apps.bea.gov/regional/histdata/releases/0718gdpstate/index.cfm
gdp <- read_csv('SQGDP2.csv')
gdp <- subset(gdp, gdp$IndustryId == 1)
colnames(gdp)[2] <- "statename"
ImExCN <- left_join(ImExCN, gdp, by = 'statename')
ImExCN <- subset(ImExCN, !is.na(ImExCN$Unit))
ImExCN <- subset(ImExCN, ImExCN$statename != 'District of Columbia')
ImExCN$share_gdp <- 100*(ImExCN$IMval2017 + ImExCN$EXval2017)/(as.numeric(ImExCN$`2017:Q4`)/1000)
library(ggrepel)

# Plot trade balance per state with coloring and size aesthetics
# Create non-aesthetic balanced trade line (dashed turquoise)
# Employ geom_jitter() and geom_abline()
# Adjust scaling for both x and y axis, scale dots
ImExCN[order(-ImExCN$share_gdp),] %>%
ggplot(aes(x = IMval2017, y = EXval2017,
  fill = as.factor(trd_main), size = share_gdp)) +
  geom_jitter(pch=21) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', color = '#00FDFF') +
  guides(size=guide_legend(title="% 2017 State GDP")) +
  guides(fill=guide_legend(title="China top partner?")) +
  coord_trans(x="log2") +
  ylim(0,20) +
  theme(legend.position = c(.15, .60)) +
  scale_fill_manual(values = c('grey', '#CC2414')) +
  geom_text_repel(aes(label=ifelse(share_gdp > 3,as.character(statename),'')),hjust=.5, vjust=-2, size = 3, color = "grey30") +
  annotate("text", x = 9.25, y = 12.25, angle = 60, label = "Balanced-Trade Line", color = "grey30", size = 5, family = "Saira Condensed") +
  scale_radius(range = c(1,12)) +
  labs(x = 'LOG IMPORTS (BILLIONS $USD)', y = 'EXPORTS (BILLIONS $USD)',
       title = 'The states rely heavily on Chinese market access for imports and exports',
       subtitle = 'In 2017, a plurality of states counted China a primary trade partner, with imports and exports representing four to \neight percent of state gross domestic product (GDP). The balanced-trade line indicates very few states run a surplus.', 
       caption = 'Data Source: United States Census Bureau & Bureau of Economic Analysis')
```

## What does a trade war entail for these states?

Microeconomic theory can be of some help to us here in describing the fundamental mechanisms operating during a trade war. Domestic businesses are usually price-takers from a global market, which with China's help has made many goods cheaper. As a result, more consumers have been able to buy cheaper goods. When a tariff is introduced, prices go up and quantity goes down, hurting the consumer by increasings costs and threfore reducing the number of people who may buy certian goods. Meanwhile, domestic suppliers can become more competitive, and can begin to reap the benefits of higher consumer costs. However, we lose some potential surplus in dead weight losses, which ends up hurting consumers, producers, and the government in the long run.

``` {r fig.width = 9, warning = F, message = F, dpi=300}
# credit to Andrew Heiss for creating this package: https://github.com/andrewheiss/reconPlots
library(devtools)
library(reconPlots)

line1 <- data.frame(Hmisc::bezier(c(10, 80, 90), c(5, 30, 45)))
line2 <- data.frame(Hmisc::bezier(c(10, 30, 90), c(45, 20, 5)))
line3 <- data.frame(Hmisc::bezier(c(0, 100), c(8, 8)))
line4 <- data.frame(Hmisc::bezier(c(0, 100), c(15, 15)))

line_intersect <- as.data.frame(curve_intersect(line1, line2))
base_intersect1 <- as.data.frame(curve_intersect(line1, line3))
base_intersect2 <- as.data.frame(curve_intersect(line2, line3))
new_intersect1 <- as.data.frame(curve_intersect(line1, line4))
new_intersect2 <- as.data.frame(curve_intersect(line2, line4))

ggplot(mapping = aes(x = x, y = y)) +
  geom_path(data = line1, color = staples_cols("red1"), size = 1) +
  geom_path(data = line2, color = staples_cols("blue2"), size = 1) +
  geom_path(data = line3, color = "black", linetype = 'dashed', size = 1) +
  geom_path(data = line4, color = "black", size = 1) +
  geom_point(data = base_intersect1, size = 2, color = 'grey40') +
  geom_point(data = base_intersect2, size = 2, color = 'grey40') +
  geom_point(data = new_intersect1, size = 2, color = 'black') +
  geom_point(data = new_intersect2, size = 2, color = 'black') +
  geom_ribbon(data = subset(line2, line2$x <= 57.33401),
      aes(ymin = 15, ymax = y), fill = staples_cols("blue1"), alpha = .2) +
  geom_ribbon(data = subset(line1, line1$x <= 36.60526),
      aes(ymin = y, ymax = 15), fill = staples_cols("red2"), alpha = .2) +
  geom_segment(data = new_intersect1,
      aes(x = x, y = 0, xend = x, yend = y), lty = "dotted") +
  geom_segment(data = new_intersect2,
      aes(x = x, y = 0, xend = x, yend = y), lty = "dotted") +  coord_equal() +
  annotate("text", x = 88, y = 17, label = 'World Price + Tariffs', family = "Saira Condensed", size = 6, color = 'black') +
  annotate("text", x = 70, y = 33, angle = 30, label = 'Domestic Supply', family = "Saira Condensed", size = 6, color = staples_cols("red1")) +
  annotate("text", x = 30, y = 31, angle = -31, label = 'Demand', family = "Saira Condensed", size = 6, color = staples_cols("blue2")) +
  annotate("text", x = 47, y = 12, label = 'Lost Consumer Surplus', family = "Saira Condensed", color = 'black', size = 4) +
  annotate("text", x = 92, y = 10, label = 'World Price', family = "Saira Condensed", size = 6, color = 'black') +
  annotate("text", x = 21, y = 22.5, label = 'New Consumer Surplus', family = "Saira Condensed", color = staples_cols("blue2"), size = 4) +
  annotate("text", x = 18, y = 12, label = 'Producer Surplus', family = "Saira Condensed", color = staples_cols("red1"), size = 4) +
  labs(x = 'QUANTITY', y = 'PRICE',
       title = 'Theory suggests tariffs hurt consumers but may help domestic suppliers',
       subtitle = 'With a given imported good, consumers are benefiting from low costs and high quantity. When tariffs push up the outside \nprice, quantity goes down and prices go up, hurting consumers while providing relatively little benefit to domestic businesses.', 
       caption = 'Credit where credit is due: Andrew Heiss for his relevant methodology')
```

## So, prices are going up... but which ones?

General equilibrium theory suggests that prices for consumer goods will go up. However, the exact goods impacted by a trade war depend on both the relative proportion of Chinese goods making up American consumption in each category, as well as the specific tariff policy of the U.S. administration. The vast majority of Chinese imports to the U.S. are in consumer electronics and appliances, trade in which totalled over $150 billion in 2017. This includes nearly all small electronics and household appliances, as well as the parts imported to produce motor vehicles by U.S. manufacturers. The three rounds of 25 percent tariffs set by the Trump administration impact many of the largest import groups, including applances, motor vehicles, and furniture. Many of the categories subject to no current tariff in the figure below have been threatened, which might also play a role in speculative price increases by corporations.

``` {r fig.width = 9, warning = F, message = F, dpi=300}
# source: https://comtrade.un.org/data/
com <- read_csv('comtrade.csv')

# Remove re-import/export data
com <- subset(com, com$`Trade Flow Code` < 3 & com$`Commodity Code` < 98)

# turn to billions USD
com_fam <- ddply(com, .(`Trade Flow`, `Commodity Code`), summarize, `Trade Value` = sum(`Trade Value (US$)`))
com_fam$`Trade Value` <- as.numeric(com_fam$`Trade Value`)
com_fam$`Trade Value` <- com_fam$`Trade Value`/1000000000

# filter to only top 25 of exports/imports
com_fam_ex <- subset(com_fam, com_fam$`Trade Flow` == 'Export')
com_fam_im <- subset(com_fam, com_fam$`Trade Flow` == 'Import')

com_fam_ex <- com_fam_ex[order(com_fam_ex$`Trade Value`),]
com_fam_im <- com_fam_im[order(com_fam_im$`Trade Value`),]

com_fam_ex$`Commodity Code` <- factor(com_fam_ex$`Commodity Code`,
                          levels = com_fam_ex$`Commodity Code`[
                          order(com_fam_ex$`Trade Value`)])
com_fam_ex <- com_fam_ex[72:96,]

com_fam_im$`Commodity Code` <- factor(com_fam_im$`Commodity Code`,
                          levels = com_fam_im$`Commodity Code`[
                          order(com_fam_im$`Trade Value`)])
com_fam_im <- com_fam_im[72:96,]

# rename, since categories are just horrible
com_fam_ex$Commodity <- ifelse(com_fam_ex$`Commodity Code` == 88, 'Aircraft & Spacecraft',
                        ifelse(com_fam_ex$`Commodity Code` == 84, 'Mechanical Machinery',
                        ifelse(com_fam_ex$`Commodity Code` == 12, 'Oils, Grains, & Straw',
                        ifelse(com_fam_ex$`Commodity Code` == 87, 'Cars, Vehicles, & Parts',
                        ifelse(com_fam_ex$`Commodity Code` == 85, 'Electronics & Equipment',
                        ifelse(com_fam_ex$`Commodity Code` == 90, 'Medical/Scientific Instruments',
                        ifelse(com_fam_ex$`Commodity Code` == 27, 'Mineral Fuels & Products',
                        ifelse(com_fam_ex$`Commodity Code` == 39, 'Plastic Goods',
                        ifelse(com_fam_ex$`Commodity Code` == 47, 'Recycled Wood Pulp & Paper',
                        ifelse(com_fam_ex$`Commodity Code` == 44, 'Wood & Charcoal',
                        ifelse(com_fam_ex$`Commodity Code` == 29, 'Organic Chemicals',
                        ifelse(com_fam_ex$`Commodity Code` == 30, 'Pharmaceuticals',
                        ifelse(com_fam_ex$`Commodity Code` == 38, 'Chemical Products',
                        ifelse(com_fam_ex$`Commodity Code` == 74, 'Copper',
                        ifelse(com_fam_ex$`Commodity Code` == 76, 'Aluminum',
                        ifelse(com_fam_ex$`Commodity Code` == 10, 'Grains & Cereals',
                        ifelse(com_fam_ex$`Commodity Code` == 3, 'Fish & Other Seafoods',
                        ifelse(com_fam_ex$`Commodity Code` == 71, 'Precious Stones & Metals',
                        ifelse(com_fam_ex$`Commodity Code` == 41, 'Hides & Leather',
                        ifelse(com_fam_ex$`Commodity Code` == 26, 'Ores & Ash',
                        ifelse(com_fam_ex$`Commodity Code` == 72, 'Iron & Steel, Raw',
                        ifelse(com_fam_ex$`Commodity Code` == 52, 'Cotton',
                        ifelse(com_fam_ex$`Commodity Code` == 28, 'Inorgnic Compounds & Rare Metals',
                        ifelse(com_fam_ex$`Commodity Code` == 48, 'Paper & Paperboard',
                        ifelse(com_fam_ex$`Commodity Code` == 73, 'Iron or Steel Goods',
                        NA)))))))))))))))))))))))))

com_fam_im$Commodity <- ifelse(com_fam_im$`Commodity Code` == 94, 'Furniture, Bedding, & Lights',
                        ifelse(com_fam_im$`Commodity Code` == 84, 'Mechanical Machinery',
                        ifelse(com_fam_im$`Commodity Code` == 95, 'Toys & Sports Equipment',
                        ifelse(com_fam_im$`Commodity Code` == 87, 'Cars, Vehicles, & Parts',
                        ifelse(com_fam_im$`Commodity Code` == 85, 'Electronics & Equipment',
                        ifelse(com_fam_im$`Commodity Code` == 90, 'Medical/Scientific Instruments',
                        ifelse(com_fam_im$`Commodity Code` == 61, 'Knitted Clothing/Apparel',
                        ifelse(com_fam_im$`Commodity Code` == 39, 'Plastic Goods',
                        ifelse(com_fam_im$`Commodity Code` == 64, 'Footwear',
                        ifelse(com_fam_im$`Commodity Code` == 44, 'Wood & Charcoal',
                        ifelse(com_fam_im$`Commodity Code` == 29, 'Organic Chemicals',
                        ifelse(com_fam_im$`Commodity Code` == 62, 'Non-Knitted Clothing/Apparel',
                        ifelse(com_fam_im$`Commodity Code` == 63, 'Textiles & Worn Clothing',
                        ifelse(com_fam_im$`Commodity Code` == 42, 'Leather Goods',
                        ifelse(com_fam_im$`Commodity Code` == 76, 'Aluminum',
                        ifelse(com_fam_im$`Commodity Code` == 83, 'Base Metal Products',
                        ifelse(com_fam_im$`Commodity Code` == 82, 'Tools & Cutlery',
                        ifelse(com_fam_im$`Commodity Code` == 71, 'Precious Stones & Metals',
                        ifelse(com_fam_im$`Commodity Code` == 40, 'Rubber Products',
                        ifelse(com_fam_im$`Commodity Code` == 96, 'Other Manufactured Goods',
                        ifelse(com_fam_im$`Commodity Code` == 70, 'Glass & Glassware',
                        ifelse(com_fam_im$`Commodity Code` == 69, 'Ceramic Products',
                        ifelse(com_fam_im$`Commodity Code` == 49, 'Printed Books & Other Print',
                        ifelse(com_fam_im$`Commodity Code` == 48, 'Paper & Paperboard',
                        ifelse(com_fam_im$`Commodity Code` == 73, 'Iron or Steel Goods',
                        NA))))))))))))))))))))))))) 

com_fam_ex$`Commodity` <- factor(com_fam_ex$`Commodity`,
                                 levels = com_fam_ex$`Commodity`[
                                   order(com_fam_ex$`Trade Value`)])
com_fam_ex <- com_fam_ex[10:25,]
com_fam_ex$tariff <- ifelse(com_fam_ex$`Commodity Code` %in% c(3, 10, 12, 52, 87, 28, 29, 30, 38), 'Tariff in Place',
                                   'No Tariff Set')

com_fam_im$`Commodity` <- factor(com_fam_im$`Commodity`,
                                 levels = com_fam_im$`Commodity`[
                                   order(com_fam_im$`Trade Value`)])
com_fam_im <- com_fam_im[10:25,]
com_fam_im$tariff <- ifelse(com_fam_im$`Commodity Code` %in% c(44, 83, 42, 29, 94), 'Round 3 (Sept 2018)',
                     ifelse(com_fam_im$`Commodity Code` %in% c(73, 39), 'Round 2 (Aug 2018)', 
                     ifelse(com_fam_im$`Commodity Code` %in% c(90, 87, 84, 85), 'Round 1 (Jul 2018)', 
                     'No Tariff Yet Set')))

ggplot(com_fam_im, aes(`Commodity`, `Trade Value`)) +
  geom_col(aes(fill = tariff), color = 'black', width = 0.8) +
  coord_flip() +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(),
        legend.position = c(.7, .40)) +
  scale_y_continuous(labels = comma, breaks = seq(0, 150, 50), expand = c(0,0)) +
  scale_fill_manual("Tariffs Set by US", values = c('light grey', '#317E99', '#149DCC', '#00FFB5')) +
  labs(x = '', y = 'TRADE IMPORTS (BILLIONS $USD)',
       title = 'Most US imports from China are consumer electronics targeted by current tariffs',
       subtitle = 'As of 2017, most Chinese imports were modern consumer staples, including electronics and house \nappliances. The Trump administration has imposed 25 percent tariffs on most of these goods.', 
       caption = 'Data Source: UN Comtrade International Trade Statistics & Personally Compiled Tariffs Listings')
```

Preliminary trade data show that these price increases have already begun, despite there being temporary holds on full tariff enactments until March 2018. The red hues on the graph below signify percentage price increases month to month, whereas blue hues signify prices falling for each good. As the horizon chart for toys shows, non-tariffed goods have followed a downward trend in price since 2016. In stark comparison are now-tariffed goods, which show sharp price increases beginning in 2018, in-line with the speculation and enactment of the tariffs in question. Most notable is the trend for appliances. Even in cases where prices have not increases, this general downward trend has stagnated. Multiple goods have returned to early 2017 levels. (THESE GRAPHS WILL BE AUGMENTED TO DATA AS THEY CONTINUE TO BE RELEASED. ALSO, I STILL NEED TO FIGURE OUT HOW TO PRESENT WITH A LEGEND, ALEX ANY ADVICE?)

``` {r fig.width = 9, fig.height = 6, warning = F, message = F, dpi=300}

# Data source: https://beta.bls.gov/dataViewer/view/8de38039b69045d6ab7fe131c44312cb
# Bureau of Labor Statistics Consumer Price Index Series
bls <- read_csv('bls_cpi.csv')
bls$new_date <- as.Date(as.yearmon(bls$date))

# Create prct diff from base date (Jan 2016) to control for initial difference in CPI
bls$prct_base <- (bls$cpi/bls$base - 1)*100

# Create rolling prct change from last 3 months
bls$prct_chng <- (bls$cpi/lag(bls$cpi, n=3) - 1)*100
bls$prct_chng <- ifelse(bls$date == 'Jan 2016' | bls$date == 'Feb 2016' | bls$date == 'Mar 2016', NA, bls$prct_chng)

# rename, since categories are just horrible
bls$id <- ifelse(bls$id == 'CUSR0000SA311', 'Clothing & Apparel',
          ifelse(bls$id == 'CUSR0000SAD', 'All Durable Goods',
          ifelse(bls$id == 'CUSR0000SAH3', 'Furniture & Bedding',
          ifelse(bls$id == 'CUSR0000SEAE', 'All Footwear',
          ifelse(bls$id == 'CUSR0000SEHK', 'Appliances',
          ifelse(bls$id == 'CUSR0000SEHM', 'Tools & Hardware',
          ifelse(bls$id == 'CUSR0000SERC', 'Sporting Goods',
          ifelse(bls$id == 'CUSR0000SERE01', 'Toys',
          ifelse(bls$id == 'CUSR0000SETA', 'Motor Vehicles',
          ifelse(bls$id == 'CUSR0000SETC', 'Vehicle Parts & Equipment',
            bls$id))))))))))

# Set the bands for horizon plot
max.y <- max(abs(bls$prct_chng), na.rm=T)
nbands = 3
hrzn_scale <- max(max.y/nbands)
h1 <- hrzn_scale
h2 <- hrzn_scale*2

bls$ypos1 <- ifelse(bls$prct_chng > 0, pmin(bls$prct_chng, h1), 0)
bls$ypos2 <- ifelse(bls$prct_chng > h1, pmin(bls$prct_chng-h1, h1), 0)
bls$ypos3 <- ifelse(bls$prct_chng > h2, pmin(bls$prct_chng-h2, h1), 0)
bls$yneg1 <- ifelse(bls$prct_chng < 0, pmin(-bls$prct_chng, h1), 0)
bls$yneg2 <- ifelse(bls$prct_chng < -h1, pmin(-bls$prct_chng-h1, h1), 0)
bls$yneg3 <- ifelse(bls$prct_chng < -h2, pmin(-bls$prct_chng-h2, h1), 0)

# Divide into explanatory graph and full graph
bls <- subset(bls, !is.na(bls$prct_chng))
bls_dur <- subset(bls, bls$id == "All Durable Goods")
bls_all <- subset(bls, bls$id != "All Durable Goods")

# Create graph of all goods in top 9 categories of imports
ggplot(bls, aes(x = new_date)) +
  geom_ribbon(aes(ymin = 0, ymax = ypos1), fill='#fecc5c', alpha=0.25) +
  geom_ribbon(aes(ymin = 0, ymax = ypos2), fill='#FF5A4C', alpha=0.25) +
  geom_ribbon(aes(ymin = 0, ymax = ypos3), fill='#CC0032', alpha=0.25) +
  geom_ribbon(aes(ymin = 0, ymax = yneg1), fill='#00FFB5', alpha=0.25) +
  geom_ribbon(aes(ymin = 0, ymax = yneg2), fill='#149DCC', alpha=0.25) +
  facet_grid(id ~ .) +
  theme(strip.text.y = element_text(angle = 0)) +
  labs(x = 'YEAR', y = 'ROLLING % PRICE CHANGE (3 MONTHS)',
       title = 'Rolling percent change in consumer prices show tariffs already impacting goods',
       subtitle = 'As of December 2018, many tariffed imports ticked up in price (red hues), particularly appliances. This includes \neverything from washing machines to iphones. Prior to the trade war, goods prices were generally falling (blue hue).', 
       caption = 'Data Sources: Bureau of Labor Statistics Consumer Price Index Series')

```

## Meanwhile, U.S. industries also take a hit

U.S. consumers aren't the only ones experiencing preliminarily negative consequences from the trade war. Producers in the United States now pay higher costs for raw materials, particuarly steel, some of which they are unable or unwilling to pass on to consumers. But in addition to this, Chinese tariffs on their exports are likely to take their toll. China is still the largest importer of U.S. goods, and as a result of the trade war several large categories of exports have had 25 percent tariffs levied against them. Those goods most impacted come from rural and old industrial communities, often the most sensitive to price shifts when it comes to laying off workers and outsourcing labor.

``` {r fig.width = 9, warning = F, message = F, dpi=300}
# Plot comparative trade in commodity groups nationally
ggplot(com_fam_ex, aes(`Commodity`, `Trade Value`)) +
  geom_col(aes(fill = tariff), color = 'black', width = 0.8) +
  coord_flip() +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(),
        legend.position = c(.7, .40)) +
  scale_y_continuous(labels = comma, breaks = seq(0, 15, 5), expand = c(0,0)) +
  scale_fill_manual("Tariffs Set by China", values = c('light grey', '#CC0032')) +
  labs(x = '', y = 'TRADE EXPORTS (BILLIONS $USD)',
       title = 'US exports to China come from the rust belt and rural industries',
       subtitle = 'As of 2017, US exports to China mostly consist of vehicles and agricultural products - both primarily \nfrom rural and midwest industrial hubs. These regions have been targeted by Chinese tariffs.', 
       caption = 'Data Source: UN Comtrade International Trade Statistics Database')
```

An analysis of exports by county finds that industry relying on Chinese markets is much more dispersed than one might first believe. While there exist hubs in Washington state, California, and New York, there are also large export areas in Texas, Georgia, Tennessee, and the Rust Belt states. This suggests tariffs levied will impact producers in several states adversly, even as they benefit from higher consumer prices.

``` {r fig.width = 9, warning = F, message = F, dpi=300}
library(maps)
library(mapdata)
library(rgdal)
library(rgeos)
library(zipcode)
library(albersusa)

# Extract zipcode specific data
data(zipcode)
zipcode$ZIP3 <- substr(zipcode$zip, 1, 3)
zipcode$ZIP3 <- as.numeric(zipcode$ZIP3)

# Source: http://tse.export.gov/metro/SelectReports.aspx?DATA=Metro
# International Trade Administration, Department of Commerce
exp <- read_csv('2017_exports_zip.csv')
exp$ZIP3 <- as.numeric(exp$ZIP3)
exp$`Full Year` <- ifelse(exp$`Full Year` == 'D', 0, exp$`Full Year`)

zip <- merge(zipcode, exp, by = 'ZIP3')
zip$`Full Year` <- as.numeric(zip$`Full Year`)
zip$zip <- as.numeric(zip$zip)

div.zip <- zip
div.zip$cnt <- 1
div.zip <- ddply(div.zip, .(ZIP3), summarize, div = sum(cnt))

# Join zip3-specific data to each zip, creating div variable
zip <- left_join(zip, div.zip, by = 'ZIP3')

# Source: http://mcdc.missouri.edu/applications/geocorr2014.html
# Missouri Census Data Center
geo <- read_csv('geocorr2014.csv')
geo$zip <- as.numeric(as.character(geo$zcta5))

# Collect population data to weight zip breakdown by
df.zip <- merge(zip, geo, by = 'zip')
df.zip$pop10 <- as.numeric(df.zip$pop10)
df.zip$afact <- as.numeric(df.zip$afact)
df.zip$`Full Year` <- as.numeric(df.zip$`Full Year`)
df.zip <- ddply(df.zip, .(county), summarize, pop = sum(pop10*afact), exports = sum(`Full Year`*afact/div))

cty_sf <- counties_sf("aeqd")
cty_sf$county <- cty_sf$fips

# Join and accumulate by county level
data.zip <- left_join(cty_sf, df.zip, by = "county")

data.zip$exp_pop <- data.zip$exports/data.zip$pop

data.zip$bins <- cut(data.zip$exp_pop, c(0,1500, 3000, 4500, 6000, 1000000), include.lowest = T, labels = c("0-1,500", "1,500-3,000", "3,000-4,500", "4,500-6,000", "6,000+"))

filter(data.zip, !is.na(bins)) %>% 
ggplot(aes(fill = bins, color = bins)) +
  geom_sf(alpha=0.8) +
  scale_fill_staples(discrete = T, palette = 'main') +
  scale_color_staples(discrete = T, palette = 'main') +
  theme(
    axis.ticks = element_line(color = NA),
    axis.line = element_line(color = NA),
    axis.text = element_text(color = NA)) +
  guides(fill=guide_legend(title="Exports per Capita ($USD)"),
         color=guide_legend(title="Exports per Capita ($USD)")) +
  labs(title = 'U.S. exports per capita are highest in the Midwest and Texas',
       subtitle = 'As of 2017, every county traded with China, with industrial hubs in Washington state, Texas, and the Rust Belt having \nhight export-per-person ratios. But this isnt unique to cities - the rural Midwest also depends on the free flow of trade.',
       caption = 'Data Sources: International Trade Administration & Missouri Census Data Center')

```

It is also important to note that, contrary to the popular narrative, rural populations are just as susceptible to interruptions of free trade with our major trade partners. An analysis of state imports and exports per capita compared to the percent of the state population in rural areas finds more rural areas may actually rely more heavily on export access to countries like China. Meanwhile, urban-centric states will likely suffer more from the increased commodity prices already cropping up for tariffed goods.

``` {r fig.width = 9, warning = F, message = F, dpi=300}
# Source: https://www.census.gov/foreign-trade/statistics/state/origin_movement/index.html#2017; https://www.census.gov/geo/reference/ua/urban-rural-2010.html
rururb <- read_csv('rur_urban2010.csv')

rururb$val_pop <- rururb$value*1000000/(rururb$POP_URBAN+rururb$POP_RURAL)
rururb$bins <- cut(rururb$POPPCT_RURAL, breaks = c(0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65),
                   include.lowest = T,
                   labels = c("0-5", "5-10", "10-15", "15-20", "20-25", "25-30", "30-35", "35-40",
                              "40-45", "45-50", "50-55", "55-60", "60+"))

rururb$val_pop <- ifelse(rururb$type == "Imports", -1*rururb$val_pop, rururb$val_pop)
mids <- ddply(rururb, .(type, bins), summarize, val_mean = mean(val_pop))

rururb <- left_join(rururb, mids, by = c('bins', 'type'))
rururb$dist <- 1- abs((rururb$val_mean - rururb$val_pop)/rururb$val_mean)

ggplot(rururb, aes(x = bins, y = val_pop, fill = type, color = type, alpha = dist)) +
  geom_point(data = subset(rururb, type = 'Imports'), stat = 'identity', size = 3) +
  geom_point(data = subset(rururb, type = 'Exports'), stat = "identity", size = 3) +
  geom_hline(yintercept = 0) +
  scale_alpha_continuous(range = c(0, 1)) +
  scale_fill_staples(discrete = T, palette = 'light') +
  guides(fill=guide_legend(title="Trade Flow"), color=guide_legend(title="Trade Flow")) +
  coord_flip() +
  scale_alpha(guide = 'none') +
  labs(x="% RURAL POPULATION", y = "VALUE PER CAPITA ($USD)",
       title = 'Rural states have just as much (or even more) to lose from export tariffs',
       subtitle = '2010 rural population data suggest a positive relationship between rural population share and export value per \ncapita. Meanwhile, large urban states consume more imports per capita, making them sensitive to import tariffs.',
       caption = 'Data Sources: International Trade Administration & U.S. Census Bureau')

```

Following this trend, the next cartogram weights each county's size by the number of farms in the state. This analysis finds acute populations of farmers across the Midwest, but also in more eastern regions and the state of Texas. These populations were among the first to be impacted by Chinese tariffs on soybeans and other foodstuffs, requiring intervention from the U.S. Department of Agriculture.

``` {r fig.width = 9, warning = F, message = F, dpi=300}

library(cartogram)
library(tmap)
library(rgdal)
# library(urbnmapr)

# Source: https://www.icip.iastate.edu/tables/agriculture/farms-by-state
farms <- read_csv('us_farms.csv')
farms$farms <- ifelse(is.na(farms$farms), 0, farms$farms)
farms$state <- farms$statename
farms$Farms <- farms$farms/farms$county_count

#---------------------------------------
#states <- get_urbn_map(map = 'states') %>%
#  rowwise() %>%
#  mutate(geometry = st_geometry(st_point(c(long, lat)))) %>%
#  select(-lat, -long) %>%
#  ungroup() %>%
#  as.data.frame() %>%
#  st_as_sf(sf_column_name = 'geometry')
#data.state <- left_join(states, farms, by = 'state_name'))
# -----------
  
state_sf <- counties_sf("aeqd")
state_sf$fips <- as.factor(as.numeric(cty_sf$state_fips))
data.state <- left_join(state_sf, farms, by = "state")

  
state.cont <- cartogram_cont(data.state, "farms", itermax = 5)

tm_shape(state.cont) + tm_polygons(col = "farms", alpha = .7, style = "jenks", legend.show = T, legend.hist = T, title = "# of Farms", palette = 'viridis', legend.position = 'right', border.col = "white") + tm_layout(frame = F, legend.outside = T, main.title = "Most U.S. farms are in the Midwest and Texas", main.title.fontfamily = "Saira Condensed", fontfamily = "Archivo Narrow", title = "Data Source: Iowa Community Indicators Program")
```

All of this being said, it is still important to note that China is not the only player impactind the U.S.-World deficit. Deficits with a variety of other nations account for the greatest share of our total national deficit. We rely on China for a fifth of our total imports, meaning with tariffs it will be difficult for other major trade partners Mexico and Canada to pick up the slack.

``` {r fig.width = 9, warning = F, message = F, dpi=300}
# Source: https://www.census.gov/foreign-trade/balance/index.html
# Use only 2017 numbers, make base unit 10s of billions
waf <- USuni[USuni$year == 2017,] %>%
        ddply(.(cty_name), summarize,
          imports = sum(imports)/10000,
          exports = sum(exports)/10000)

# Get top 5 trade partners for manual input
# TO DO: FIGURE OUT [[]] AUTOMATED METHOD
waf_imp <- waf[order(-waf$imports),][1:5,]
waf_exp <- waf[order(-waf$exports),][1:5,]

imp <- c(`China` = 50,
         `Mexico` = 31,
         `Canada` = 30,
         `Japan` = 14,
         `Germany` = 12,
         `All Others` = 97)

exp <- c(`China` = 13,
         `Mexico` = 24,
         `Canada` = 28,
         `Japan` = 7,
         `Germany` = 5,
         `All Others` = 77,
         80)

library(waffle)

par(family = "Saira Condensed")
iron(
  waffle(exp, rows = 8, size = 1,
         colors = c("#CC0032", "#FF5A4C", "#00FFB5", "#149DCC", "#317E99", 'grey70', 'white'),
         title = 'U.S. Exports to the World',
         xlab = '1 square = $10 billion in trade'),
  waffle(imp, rows = 8, size = 1,
         colors = c("#CC0032", "#FF5A4C", "#00FFB5", "#149DCC", "#317E99", 'grey70'),
         title = 'U.S. Imports from the World',
         xlab = 'Data Source: United States Census Bureau')
  )

```