---
title: "Who's the Victim in a US-China Trade War?"
author: "Authored by Stephen Stapleton"
date: "January 27, 2019"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(zoo)
library(dplyr)
library(here)
library(readxl)
library(ggrepel)
library(scales)
```

## Introduction

Trade wars tend to captivate the media and the broader public, with flights of coverage going through the chaotic highs and lows of retaliatory and conciliatory actions. In such moments, it is often difficult to decouple the national drama from the facts of what a trade war truly is, and, perhaps more importantly, who is left paying the tab when a new trade equilibrium is reached. This project is intended to provide a short respite from the back and forth, and to go into the specifics of what a trade war means for the American people. My mission is to go deep into the numbers, and to give some of the best estimates as to what changes in the global market we may expect, and what regions are likely to be hit hardest with job layoffs and rising costs.

## First, why China?

The question is a legitimate one, and is the best place to start to give an idea as to why this trade war began. When the Chinese market first opened up in earnest to the world, it provided an opportunity for many companies looking for cheap labor to make a substantial profit by producing cheap and selling at normal prices to developed nation consumers. Since then, trade between China and the World has exploded, lifting billions out of poverty in China and causing substantial drops in the price of many consumer goods in the United States. This has been a net contributor to American households, making many consumer electronics attainable for even lower-income households. The United State runs the largest trade deficit in the World with China, since as a nation we buy substantially more than we sell to them. Part of this growing trend was China's entry into the World Trade Organization in 2001, which eliminated many of the trade barriers that restricted free trade. While weighting our trade balances by gross domestic product (GDP) indicates this imbalance is higher per capita in many other nations, policymakers nevertheless have expressed concern that without intervention, this imbalance will continue to widen.

```{r fig.width = 9, warning = F, message = F}
# Source: https://www.census.gov/foreign-trade/balance/index.html
USall <- read_csv('country.csv')

# Parse dates for use in ggplot
USall$new_date <- as.Date(as.yearmon(USall$date))

# Redefine variable deficit to trade balance
USall$trd_bal <- -USall$deficit

# Restrict to years 1985-2017 for yearly total
USall <- USall[USall$new_date < '2018-09-01',]

# Exclude regional trade comparisons
USuni <- USall[USall$cty_code > 1000,]

# Summarize over year to adjust for seasonal spikes
USyr <- USuni %>%
  group_by(cty_code, year) %>%
  summarize(trd_bal = sum(trd_bal))

# Prepare for ggplot
USyr$is_china <- ifelse(USyr$cty_code == 5700, 1, 0)
USyr$trd_bal <- USyr$trd_bal/1000
USyr <- USyr[USyr$year < 2018,]

# Plot trade balance per country with coloring and alpha aesthetics
# Create non-aesthetic xintercept for WTO adoption by China
# Employ geom_line(), geom_vline()
# Adjust scaling for both x and y axis
ggplot(USyr, aes(x = year, y = trd_bal,
  color = as.factor(cty_code), alpha = as.factor(is_china))) +
  geom_line() +
  geom_vline(xintercept = 2001, linetype = 'dashed', color = 'turquoise') +
  scale_alpha_manual(values = c(.4, 1)) +
  scale_color_manual(values = c(rep('grey', 150), 'red', rep('grey', 93))) +
  scale_x_continuous(breaks = seq(1985, 2017, 5), expand = c(0,0)) +
  scale_y_continuous(breaks = seq(-400, 100, 100), expand = c(0,0)) +
  theme_classic() +
  annotate("text", x = 2015, y = -300, label = 'China', color = 'red') +
  annotate("text", x = 2000.5, y = -200, label = 'China enters WTO', angle = 90) +
  annotate("rect", xmin = 2008, xmax = 2009, ymin = -400, ymax = 100, alpha = .1) +
  theme(legend.position = 'none') +
  labs(x = 'Year', y = 'Trade Balance (billions $USD)',
       title = 'The US-China trade gap stands out among US trade relationships',
       subtitle = 'In nominal terms, the deficit trends downward after China entered the World Trade Organization in December 2001, \nquadrupling since 2000. While population weighting reduces the deficit somewhat, it does not change this trend.', 
       caption = 'Data Source: United States Census Bureau')
```

Breaking the trade relationship down to the state level, China runs a surplus with many of its primary state trading partners. A balance-trade line shows most states with China as their primary foreign trade partner run a deficit, with Washington State a notable outlier. Many states rely substantially on trade with China as a part of their local economy; California, Illinois, Washington, and Georgia each attribute roughly five to eight percent of their state GDP to trade with China (imports and exports). These states are then ones with particularly more to lose from a US-China trade war, and will be explored in greater detail as this project continues. 

``` {r fig.width = 9, warning = F, message = F}
# source: https://www.census.gov/foreign-trade/statistics/state/data/index.html
EXall <- read_excel('exctyall.xls',
            range = cell_rows(5:1436),
            col_types = c(rep('text', 3), rep('numeric', 9))
            )
IMall <- read_excel('imctyall.xls',
            range = cell_rows(5:1436),
            col_types = c(rep('text', 3), rep('numeric', 9))
            )

# extract only 2017 imports/exports
IMall$IMval2017 <- IMall$val2017
EXall$EXval2017 <- EXall$val2017

IMall$IMshr2017 <- IMall$share17
EXall$EXshr2017 <- EXall$share17

IMall$IMrank <- IMall$rank
EXall$EXrank <- EXall$rank

# combine by state and trading country
ImEx <- left_join(IMall, EXall, by = c('statename', 'countryd'))

# subset only chinese imports and exports
ImExCN <- subset(ImEx, (ImEx$countryd %in% 'China'))

# specify if china is top trading partner
ImExCN$trd_main <- ifelse(ImExCN$IMrank %in% c(1:3) & ImExCN$EXrank %in% c(1:3), 'Yes', 'No')

# convert millions to billions for each of graph reading
ImExCN$IMval2017 <- ImExCN$IMval2017/1000
ImExCN$EXval2017 <- ImExCN$EXval2017/1000

ImExCN$IMshr2017 <- ImExCN$IMshr2017/100
ImExCN$EXshr2017 <- ImExCN$EXshr2017/100

# Create relative measure of (imports+exports)/gdp per state
# Source: https://apps.bea.gov/regional/histdata/releases/0718gdpstate/index.cfm
gdp <- read_csv('SQGDP2.csv')
gdp <- subset(gdp, gdp$IndustryId == 1)
colnames(gdp)[2] <- "statename"
ImExCN <- left_join(ImExCN, gdp, by = 'statename')
ImExCN <- subset(ImExCN, !is.na(ImExCN$Unit))
ImExCN <- subset(ImExCN, ImExCN$statename != 'District of Columbia')
ImExCN$share_gdp <- 100*(ImExCN$IMval2017 + ImExCN$EXval2017)/(as.numeric(ImExCN$`2017:Q4`)/1000)
library(ggrepel)

# Plot trade balance per state with coloring and size aesthetics
# Create non-aesthetic balanced trade line (dashed turquoise)
# Employ geom_jitter() and geom_abline()
# Adjust scaling for both x and y axis, scale dots
ggplot(ImExCN, aes(x = IMval2017, y = EXval2017,
  fill = as.factor(trd_main), size = share_gdp)) +
  geom_jitter(pch=21) +
  theme_bw() +
  theme(axis.line = element_line(colour = "black"), panel.border = element_blank()) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', color = 'turquoise') +
  guides(size=guide_legend(title="% 2017 State GDP")) +
  guides(fill=guide_legend(title="China top partner?")) +
  coord_trans(x="log2") +
  ylim(0,20) +
  theme(legend.position = c(.15, .66), legend.background = element_rect(fill=NA)) +
  scale_fill_manual(values = c('grey', 'red')) +
  geom_text_repel(aes(label=ifelse(share_gdp > 4 & trd_main == "Yes",as.character(statename),'')),hjust=.5, vjust=-2, size = 3) +
  annotate("text", x = 9.25, y = 12.25, angle = 60, label = "Balanced-Trade Line") +
  scale_radius(range = c(1,12)) +
  labs(x = 'Log Imports (billions USD)', y = 'Exports (billions USD)',
       title = 'The states rely heavily on Chinese market access for imports and exports',
       subtitle = 'In 2017, a plurality of states counted China their primary trade partner, with these imports and exports alone representing four to \neight percent of state gross domestic product (GDP). The dotted balanced-trade line indicates very few states run a surplus.', 
       caption = 'Data Source: United States Census Bureau & Bureau of Economic Analysis')
```

## What does a trade war entail for these states?

Microeconomic theory can be of some help to us here in describing the fundamental mechanisms operating during a trade war. Domestic businesses are usually price-takers from a global market, which with China's help has made many goods cheaper. As a result, more consumers have been able to buy cheaper goods. When a tariff is introduced, prices go up and quantity goes down, which hurts the consumer but may help domestic producers be better competitive. However, we lose some potential surplus in dead weight losses, which ends up hurting consumers, producers, and the government in the long run.

``` {r fig.width = 9, warning = F, message = F}
# credit to Andrew Heiss for creating this package: https://github.com/andrewheiss/reconPlots
library(devtools)
library(reconPlots)

line1 <- data.frame(Hmisc::bezier(c(10, 80, 90), c(10, 50, 90)))
line2 <- data.frame(Hmisc::bezier(c(10, 30, 90), c(90, 30, 10)))
line3 <- data.frame(Hmisc::bezier(c(0, 100), c(15, 15)))
line4 <- data.frame(Hmisc::bezier(c(0, 100), c(25, 25)))

line_intersect <- as.data.frame(curve_intersect(line1, line2))
base_intersect1 <- as.data.frame(curve_intersect(line1, line3))
base_intersect2 <- as.data.frame(curve_intersect(line2, line3))
new_intersect1 <- as.data.frame(curve_intersect(line1, line4))
new_intersect2 <- as.data.frame(curve_intersect(line2, line4))

ggplot(mapping = aes(x = x, y = y)) +
  geom_path(data = line1, color = "red", size = 1) +
  geom_path(data = line2, color = "blue", size = 1) +
  geom_path(data = line3, color = "black", linetype = 'dashed', size = 1) +
  geom_path(data = line4, color = "black", size = 1) +
  geom_point(data = base_intersect1, size = 2) +
  geom_point(data = base_intersect2, size = 2) +
  geom_ribbon(data = subset(line2, line2$x <= 58.51075),
      aes(ymin = 25, ymax = y), fill = 'light blue', alpha = .2) +
  geom_ribbon(data = subset(line1, line1$x <= 34.13912),
      aes(ymin = y, ymax = 25), fill = 'red', alpha = .2) +
  geom_segment(data = new_intersect1,
      aes(x = x, y = 0, xend = x, yend = y), lty = "dotted") +
  geom_segment(data = new_intersect2,
      aes(x = x, y = 0, xend = x, yend = y), lty = "dotted") +  coord_equal() +
  theme_classic() +
  annotate("text", x = 85, y = 27.5, label = 'World Price + Tariffs', color = 'black') +
  annotate("text", x = 75, y = 55, angle = 50, label = 'Domestic Supply', color = 'red') +
  annotate("text", x = 30, y = 60, angle = -60, label = 'Demand', color = 'blue') +
  annotate("text", x = 47, y = 20, label = 'Lost Consumer Surplus', color = 'black', size = 3) +
  annotate("text", x = 92, y = 17.5, label = 'World Price', color = 'black') +
  annotate("text", x = 25, y = 35, label = 'Consumer Surplus', color = 'blue', size = 3) +
  labs(x = 'Quantity', y = 'Price',
       title = 'Theory suggests tariffs hurt consumers but may help domestic suppliers',
       subtitle = 'With a given imported good, consumers are benefiting from low costs and high quantity. \nWhen tariffs push up the outside price, quantity goes down and prices go up, \nhurting consumers while providing relatively little benefit to domestic businesses.', 
       caption = 'Credit where credit is due: Andrew Heiss for his relevant methodology')
```

## Which industries are at risk?

When we talk about prices and potential job losses, it is always important to say that goods are in an entire ecosystem that can adapt to new environmets. That being said, there are several industries more likely to experience price increases in the coming year. Particularly poignant are consumer electronics and other staples of the modern American. Twenty-five percent tariffs on these goods will likely be passed on to consumers.

When it comes to job loss, tariffs exacted by China are generally bad for hubs of old industry and rural production, such as the rust belt and wide swaths of the midwest. These industries will find fewer places to bring goods to market, and will likely restrict employment numbers or cut back on current employees.

``` {r fig.width = 9, warning = F, message = F }
# source: https://comtrade.un.org/data/
com <- read_csv('comtrade.csv')

# Remove re-import/export data
com <- subset(com, com$`Trade Flow Code` < 3 & com$`Commodity Code` < 98)

# turn to billions USD
com_fam <- com %>%
  group_by(`Trade Flow`, `Commodity Code`) %>%
  summarize(`Trade Value` = sum(`Trade Value (US$)`))
com_fam$`Trade Value` <- as.numeric(com_fam$`Trade Value`)
com_fam$`Trade Value` <- com_fam$`Trade Value`/1000000000

# filter to only top 25 of exports/imports
com_fam_ex <- subset(com_fam, com_fam$`Trade Flow` == 'Export')
com_fam_im <- subset(com_fam, com_fam$`Trade Flow` == 'Import')

com_fam_ex <- com_fam_ex[order(com_fam_ex$`Trade Value`),]
com_fam_im <- com_fam_im[order(com_fam_im$`Trade Value`),]

com_fam_ex$`Commodity Code` <- factor(com_fam_ex$`Commodity Code`,
                          levels = com_fam_ex$`Commodity Code`[
                          order(com_fam_ex$`Trade Value`)])
com_fam_ex <- com_fam_ex[72:96,]

com_fam_im$`Commodity Code` <- factor(com_fam_im$`Commodity Code`,
                          levels = com_fam_im$`Commodity Code`[
                          order(com_fam_im$`Trade Value`)])
com_fam_im <- com_fam_im[72:96,]

# rename, since categories are just horrible
com_fam_ex$Commodity <- ifelse(com_fam_ex$`Commodity Code` == 88, 'Aircraft & Spacecraft',
                        ifelse(com_fam_ex$`Commodity Code` == 84, 'Mechanical Machinery',
                        ifelse(com_fam_ex$`Commodity Code` == 12, 'Oils, Grains, & Straw',
                        ifelse(com_fam_ex$`Commodity Code` == 87, 'Cars, Vehicles, & Parts',
                        ifelse(com_fam_ex$`Commodity Code` == 85, 'Electronics & Equipment',
                        ifelse(com_fam_ex$`Commodity Code` == 90, 'Medical/Scientific Instruments',
                        ifelse(com_fam_ex$`Commodity Code` == 27, 'Mineral Fuels & Products',
                        ifelse(com_fam_ex$`Commodity Code` == 39, 'Plastic Goods',
                        ifelse(com_fam_ex$`Commodity Code` == 47, 'Recycled Wood Pulp & Paper',
                        ifelse(com_fam_ex$`Commodity Code` == 44, 'Wood & Charcoal',
                        ifelse(com_fam_ex$`Commodity Code` == 29, 'Organic Chemicals',
                        ifelse(com_fam_ex$`Commodity Code` == 30, 'Pharmaceuticals',
                        ifelse(com_fam_ex$`Commodity Code` == 38, 'Chemical Products',
                        ifelse(com_fam_ex$`Commodity Code` == 74, 'Copper',
                        ifelse(com_fam_ex$`Commodity Code` == 76, 'Aluminum',
                        ifelse(com_fam_ex$`Commodity Code` == 10, 'Grains & Cereals',
                        ifelse(com_fam_ex$`Commodity Code` == 3, 'Fish & Other Seafoods',
                        ifelse(com_fam_ex$`Commodity Code` == 71, 'Precious Stones & Metals',
                        ifelse(com_fam_ex$`Commodity Code` == 41, 'Hides & Leather',
                        ifelse(com_fam_ex$`Commodity Code` == 26, 'Ores & Ash',
                        ifelse(com_fam_ex$`Commodity Code` == 72, 'Iron & Steel, Raw',
                        ifelse(com_fam_ex$`Commodity Code` == 52, 'Cotton',
                        ifelse(com_fam_ex$`Commodity Code` == 28, 'Inorgnic Compounds & Rare Metals',
                        ifelse(com_fam_ex$`Commodity Code` == 48, 'Paper & Paperboard',
                        ifelse(com_fam_ex$`Commodity Code` == 73, 'Iron or Steel Goods',
                        NA)))))))))))))))))))))))))

com_fam_im$Commodity <- ifelse(com_fam_im$`Commodity Code` == 94, 'Furniture, Bedding, & Lights',
                        ifelse(com_fam_im$`Commodity Code` == 84, 'Mechanical Machinery',
                        ifelse(com_fam_im$`Commodity Code` == 95, 'Toys & Sports Equipment',
                        ifelse(com_fam_im$`Commodity Code` == 87, 'Cars, Vehicles, & Parts',
                        ifelse(com_fam_im$`Commodity Code` == 85, 'Electronics & Equipment',
                        ifelse(com_fam_im$`Commodity Code` == 90, 'Medical/Scientific Instruments',
                        ifelse(com_fam_im$`Commodity Code` == 61, 'Knitted Clothing/Apparel',
                        ifelse(com_fam_im$`Commodity Code` == 39, 'Plastic Goods',
                        ifelse(com_fam_im$`Commodity Code` == 64, 'Footwear',
                        ifelse(com_fam_im$`Commodity Code` == 44, 'Wood & Charcoal',
                        ifelse(com_fam_im$`Commodity Code` == 29, 'Organic Chemicals',
                        ifelse(com_fam_im$`Commodity Code` == 62, 'Non-Knitted Clothing/Apparel',
                        ifelse(com_fam_im$`Commodity Code` == 63, 'Textiles & Worn Clothing',
                        ifelse(com_fam_im$`Commodity Code` == 42, 'Leather Goods',
                        ifelse(com_fam_im$`Commodity Code` == 76, 'Aluminum',
                        ifelse(com_fam_im$`Commodity Code` == 83, 'Base Metal Products',
                        ifelse(com_fam_im$`Commodity Code` == 82, 'Tools & Cutlery',
                        ifelse(com_fam_im$`Commodity Code` == 71, 'Precious Stones & Metals',
                        ifelse(com_fam_im$`Commodity Code` == 40, 'Rubber Products',
                        ifelse(com_fam_im$`Commodity Code` == 96, 'Other Manufactured Goods',
                        ifelse(com_fam_im$`Commodity Code` == 70, 'Glass & Glassware',
                        ifelse(com_fam_im$`Commodity Code` == 69, 'Ceramic Products',
                        ifelse(com_fam_im$`Commodity Code` == 49, 'Printed Books & Other Print',
                        ifelse(com_fam_im$`Commodity Code` == 48, 'Paper & Paperboard',
                        ifelse(com_fam_im$`Commodity Code` == 73, 'Iron or Steel Goods',
                        NA)))))))))))))))))))))))))                                                                                                            

com_fam_ex$`Commodity` <- factor(com_fam_ex$`Commodity`,
                                 levels = com_fam_ex$`Commodity`[
                                   order(com_fam_ex$`Trade Value`)])
com_fam_ex <- com_fam_ex[10:25,]
com_fam_ex$tariff <- ifelse(com_fam_ex$`Commodity Code` %in% c(3, 10, 12, 52, 87, 28, 29, 30, 38), 'Tariff in Place',
                                   'No Tariff Set')

com_fam_im$`Commodity` <- factor(com_fam_im$`Commodity`,
                                 levels = com_fam_im$`Commodity`[
                                   order(com_fam_im$`Trade Value`)])
com_fam_im <- com_fam_im[10:25,]
com_fam_im$tariff <- ifelse(com_fam_im$`Commodity Code` %in% c(44, 83, 42, 29, 94), 'Round 3 (Sept 2018)',
                     ifelse(com_fam_im$`Commodity Code` %in% c(73, 39), 'Round 2 (Aug 2018)', 
                     ifelse(com_fam_im$`Commodity Code` %in% c(90, 87, 84, 85), 'Round 1 (Jul 2018)', 
                     'No Tariff Set')))

# Plot comparative trade in commodity groups nationally
ggplot(com_fam_ex, aes(`Commodity`, `Trade Value`)) +
  geom_col(aes(fill = tariff), color = 'black', width = 0.8) +
  coord_flip() +
  theme_bw() +
  theme(axis.line = element_line(colour = "black"), panel.border = element_blank(),
        panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(),
        legend.position = c(.7, .40), legend.background = element_rect(fill=NA)) +
  scale_y_continuous(labels = comma, breaks = seq(0, 15, 5), expand = c(0,0)) +
  scale_fill_manual("Tariffs Set by China", values = c('light grey', 'red')) +
  labs(x = '', y = 'Trade Exports (billions $USD)',
       title = 'US exports to China come from the rust belt and rural industries',
       subtitle = 'As of 2017, US exports to China mostly consist of vehicles and agricultural products - both primarily \nfrom rural and midwest industrial hubs. These regions have been targeted by Chinese tariffs.', 
       caption = 'Data Source: UN Comtrade International Trade Statistics Database')
```

``` {r fig.width = 9, warning = F, message = F }
ggplot(com_fam_im, aes(`Commodity`, `Trade Value`)) +
  geom_col(aes(fill = tariff), color = 'black', width = 0.8) +
  coord_flip() +
  theme_bw() +
  theme(axis.line = element_line(colour = "black"), panel.border = element_blank(),
        panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(),
        legend.position = c(.7, .40), legend.background = element_rect(fill=NA)) +
  scale_y_continuous(labels = comma, breaks = seq(0, 150, 50), expand = c(0,0)) +
  scale_fill_manual("Tariffs Set by US", values = c('light grey', 'dark red', 'red', 'pink')) +
  labs(x = '', y = 'Trade Imports (billions $USD)',
       title = 'Most US imports from China are consumer electronics targeted by current tariffs',
       subtitle = 'As of 2017, Chinese imports were most modern consumer staples, including electronics and household \nappliances. The Trump administration has imposed 25 percent tariffs on most of these goods.', 
       caption = 'Data Source: UN Comtrade International Trade Statistics & Personally Compiled Tariffs Listings')
```

